# Packages

{{< video https://vimeo.com/1198721978/d3ec236a6b?fl=ml&fe=ec >}}

R's power comes from its ecosystem of packages. This video shows how to install and load packages and introduces the tidyverse collection we will use throughout the course.


```{r, eval = FALSE}
# Loading a package that is installed on your machine, put on top of your script.
library(tidyverse)

# Installing other packages, don't put this in your script
install.packages("readxl")
install.packages("gapminder")
install.packages("nycflights")
install.packages("stargazer")
install.packages("broom")
install.packages("patchwork")
```

## Exercises

### Exercise 1.5.1

Load the tidyverse with `library(tidyverse)`. What packages are listed as "Attaching core tidyverse packages" in the message?

<details> <summary>Solution</summary>

```r
library(tidyverse)
```

```
── Attaching core tidyverse packages ──
✔ dplyr     1.1.4     ✔ readr     2.1.5
✔ forcats   1.0.0     ✔ stringr   1.5.1
✔ ggplot2   3.5.1     ✔ tibble    3.2.1
✔ lubridate 1.9.3     ✔ tidyr     1.3.1
✔ purrr     1.0.2
```

The core packages are: dplyr, forcats, ggplot2, lubridate, purrr, readr, stringr, tibble, and tidyr. (The exact versions will depend on when you installed.)

</details>

### Exercise 1.5.2

What happens if you try to use a function from a package you have not loaded? Run the following _without_ loading `readxl` first:

```r
read_excel("test.xlsx")
```

What error do you get? Now load the package and explain why the error would go away (you do not need an actual Excel file).

<details> <summary>Solution</summary>

```r
read_excel("test.xlsx")
```

```
Error in read_excel("test.xlsx") : could not find function "read_excel"
```

The error says R cannot find the function. This is because `read_excel` lives in the `readxl` package, which has not been loaded. After running `library(readxl)`, R knows where to find the function. (You would then get a different error about the file not existing, which is expected since we have no such file.)

</details>

### Exercise 1.5.3

Run `installed.packages()` inside of the `nrow()`-function to count how many packages are installed on your system:

```r
nrow(installed.packages())
```

<details> <summary>Solution</summary>

```{r}
nrow(installed.packages())
```

Your number wil probably be differenty. The tidyverse alone installs many other packages upon which it depends, so even a fresh R installation with tidyverse will have 100+ packages.

</details>

### Exercise 1.5.4

The function `sessionInfo()` prints information about your R session, including which packages are loaded. Run it. Can you identify the "attached" packages (those you loaded explicitly with `library()`) versus the "loaded via a namespace" packages (dependencies loaded automatically)?

<details> <summary>Solution</summary>

The output will contain a section like:

```{r}
sessionInfo()
```

"Attached" packages are those whose functions are directly available by name. "Loaded via namespace" packages are available internally (other packages can use them) but their functions are not directly on the search path — you would need to use `package::function()` to call them.

</details>

### Exercise 1.5.5

Try installing a package that does not exist:

```r
install.packages("this_package_does_not_exist_xyz")
```

What does the error/warning message say? Understanding installation error messages is useful for diagnosing real problems (typos in package names, network issues, etc.).

<details> <summary>Solution</summary>

```{r, error = TRUE, warning = TRUE}
install.packages("this_package_does_not_exist_xyz")
```

The message says the package is "not available". In practice, if you get this message for a real package, the most common causes are: (1) a typo in the name, (2) the package was removed from CRAN, or (3) a network connectivity problem.

</details>
