## Vectors

{{< video https://vimeo.com/1198721980/2478cb4a26?fl=ml&fe=ec  >}}

Vectors are the fundamental data structure in R. This video covers how to create and manipulate vectors and introduces vectorised operations.

```{r, eval = FALSE}
# Vectors
prices   <- c(29.90, 35, 11, 12.5, 48, 22.75)
product  <- c("coffee", "tea", "juice", "sandwich", "cookie")
in_stock <- c(TRUE, TRUE, FALSE, TRUE, TRUE)
quantities <- c(3, 5, 0, 2, 8)

# Arithmetics on vectors
revenue <- prices * quantities
prices * 1.1

# Applying functions to vectors
length(prices)
sum(prices)
mean(prices)
min(prices)
max(prices)
sd(prices)

# Other ways to create vectors
1:10                                       # Sequence of numbers
seq(from = 0, to = 1, by = 0.1)            # Sequence, specifying step size
seq(from = 0, to = 3.14, length.out = 25)  # Sequence, specifying length
rep(0, times = 5)                          # Repeating element
rep(c("A", "B", times = 5))                # Repeating patterns

# Logical comparisons
prices > 30
sum(prices > 30)
```

## Exercises

### Exercise 1.7.1

Create a vector called `temps` containing the values 18.5, 21.0, 16.3, 22.7, 19.8. Compute the mean and standard deviation.

<details> <summary>Solution</summary>

```{r}
temps <- c(18.5, 21.0, 16.3, 22.7, 19.8)
mean(temps)
sd(temps)
```

</details>

### Exercise 1.7.2

Create a vector of all even numbers from 2 to 20 using `seq()`. Compute the sum.

<details> <summary>Solution</summary>

```{r}
evens <- seq(from = 2, to = 20, by = 2)
evens
sum(evens)
```

</details>

### Exercise 1.7.3

Given the following vector:

```r
sales <- c(120, 85, 200, 340, 95, 180, 420, 60, 275, 150)
```

How many values are above 200? What proportion of values is above 200?

<details> <summary>Solution</summary>

```{r}
sales <- c(120, 85, 200, 340, 95, 180, 420, 60, 275, 150)

# Count
sum(sales > 200)

# Proportion
mean(sales > 200)
```

Three values (340, 420, 275) are strictly above 200, which is 30% of the 10 values. The `mean()` of a logical vector gives the proportion of `TRUE` values.

</details>

### Exercise 1.7.4

Create two vectors:

```{r}
prices <- c(50, 30, 80, 120, 45)
quantities <- c(10, 25, 5, 2, 15)
```

Compute a vector of revenues (price × quantity). What is the total revenue? Which product has the highest revenue?

<details> <summary>Solution</summary>

```{r}
revenue <- prices * quantities
revenue
sum(revenue)
which.max(revenue)
```

Total revenue is 2565. Product 2 (price 30, quantity 25) has the highest revenue at 750.

</details>

### Exercise 1.7.5

What happens when you combine different types in one vector? Predict the result, then check:

```r
mixed <- c(1, 2, "three", 4, TRUE)
mixed
class(mixed)
```

Can you explain the rule R uses?

<details> <summary>Solution</summary>

```{r}
mixed <- c(1, 2, "three", 4, TRUE)
mixed
class(mixed)
```

Everything was converted to character strings. R uses **type coercion**: a vector can only hold one type, so R converts all elements to the most general type. The hierarchy is: logical → numeric → character. Since `"three"` is a character, everything becomes character. This is an important source of bugs, for example, if a column in a CSV has one non-numeric entry, the whole column may be read as text if you are not careful.

</details>