## Variables and types

{{< video https://vimeo.com/1198721979/e6735c2289?fl=ml&fe=ec  >}}

We introduce variables as named containers for data and explore R's basic data types: numeric, character, logical, and factors.


```{r, eval = FALSE}
# Assigning variables
x <- 5

# Using variables
x * 3
y <- 10
x + y
z <- x + y

# Variable names are case sensitive
mydata <- 12
myData <- 11

# We will use snake_case for variable names. Good examples:
monthly_revenue <- 5000
tax_rate <- 0.25
company_name <- "Equinor"

# Bad names:
#   - MonthlyRevenue      <- CamelCase
#   - x                   <- Not descriptive
#   - 2nd_quarter         <- Starts with a number, causes error

# Numeric data type
price <- 99.50
quantity <- 7

# Character data type
product <- "coffee"
greeting <- "Hello, World"

# Logical data type
in_stock <- TRUE
on_sale <- FALSE

# Check data type
class(price)
class(product)
class(in_stock)

# R overwrites variables, no questions or warnings
x <- 5
x <- 10
```

## Exercises

### Exercise 1.6.1

Create three variables: `product` with the value `"laptop"`, `price` with the value `8999.50`, and `in_stock` with the value `TRUE`. Check the type of each using `class()`.

<details> <summary>Solution</summary>

```{r}
product <- "laptop"
price <- 8999.50
in_stock <- TRUE

class(product)
class(price)
class(in_stock)
```


</details>

### Exercise 1.6.2

What happens when you run the following code? Explain each line.

```r
a <- 10
b <- a
a <- 20
b
```

<details> <summary>Solution</summary>

```{r}
a <- 10   # a is assigned the value 10
b <- a    # b is assigned the current value of a, which is 10
a <- 20   # a is reassigned to 20; b is NOT affected
b         # b is still 10
```

The key insight is that `b <- a` copies the _value_ of `a` at that moment. It does *not* create a link between `a` and `b`. Changing `a` later does not change `b`. Python users, beware!

</details>

### Exercise 1.6.3

Try the following and explain the results:

```r
TRUE + TRUE
FALSE + TRUE
TRUE * 5
sum(c(TRUE, FALSE, TRUE, TRUE))
```

<details> <summary>Solution</summary>

```{r}
TRUE + TRUE
FALSE + TRUE
TRUE * 5
sum(c(TRUE, FALSE, TRUE, TRUE))
```

When used in arithmetic, R converts `TRUE` to 1 and `FALSE` to 0. This is why `sum()` of a logical vector counts the number of `TRUE` values, which is a very useful trick.

</details>

### Exercise 1.6.4

What is the result of each of the following? Before running them, try to predict the answer:

```r
class(5L)
class(5)
class("5")
class(5 > 3)
```

<details> <summary>Solution</summary>

```{r}
class(5L)
class(5)
class("5")
class(5 > 3)
```

`5L` is explicitly an integer (the `L` suffix forces it). `5` without `L` is a double-precision numeric (Rs default). `"5"` is a character string (the quotes make it text). `5 > 3` evaluates to `TRUE`, which is a logical.

</details>

### Exercise 1.6.5

What happens if you assign a value to a variable named `mean`? Can you still use the `mean()` function afterwards? Try it:

```r
mean <- 42
mean
mean(c(1, 2, 3))
```

What is going on, and why should you be careful about naming variables?

<details> <summary>Solution</summary>

```r
mean <- 42
mean
```

```
[1] 42
```

```r
mean(c(1, 2, 3))
```

```
[1] 2
```

Both work! When you type `mean` alone, R finds your variable. When you type `mean(...)` with parentheses, R recognises it as a function call and finds the built-in `mean()` function. R keeps variables and functions in slightly different places.

However, this is **dangerous and confusing**. While R can often sort it out, it makes your code much harder to read. As a rule: never name your variables after existing R functions. Avoid names like `mean`, `sum`, `data`, `df`, `c`, `t`, `list`, etc.

To clean up:

```r
rm(mean)  # remove the variable, restoring the function name
```

</details>


