## Data frames

{{< video  https://vimeo.com/1198722264/3f13cdc037?fl=ml&fe=ec >}}

Data frames are R's representation of tabular data. This video introduces how to create, inspect, and perform basic operations on data frames.

```{r, eval = FALSE}
# Creating a simple data frame
prices <- c(29.90, 35, 11, 12.5, 48, 22.75)
products <- c("coffee", "tea", "juice", "sandwich", "cookie")
quantities <- c(3, 5, 0, 2, 8)

sales <- data.frame(
  product = products,
  price = prices,
  quantity = quantities
)

# Applying functions to the data frame
head(sales)
tail(sales)
nrow(sales)
ncol(sales)
names(sales)
str(sales)
glimpse(sales) # Requires that the tidyverse package is loaded

# Subsetting data frames
sales$product
mean(sales$price)
sum(sales$quantity)

# Using double brackets instead
sales[["price"]]

# Subsetting rows and columns
sales[2, 3]
sales[1, ]
sales[, "price"]
sales[sales$quantity > 0, ]

# Using the tidyverse "tibble" instead of the built-in data frame
library(tidyverse)

sales <- tibble(
  product = products,
  price = prices,
  quantity = quantities
)

# Subsetting a tibble always returns a new tibble, regardless of dimension!
```

## Exercises

### Exercise 1.9.1

Create a tibble called `students` with three columns: `name` (character), `grade` (numeric), and `passed` (logical), containing at least four rows of made-up data.

<details> <summary>Solution</summary>

```{r, message = FALSE}
library(tidyverse)

students <- tibble(
  name = c("Anna", "Bob", "Charlie", "David"),
  grade = c(85, 62, 91, 45),
  passed = c(TRUE, TRUE, TRUE, FALSE)
)

students
```

</details>

### Exercise 1.9.2

Using your `students` tibble, compute the average grade using `$` to extract the column.

<details> <summary>Solution</summary>

```{r}
mean(students$grade)
```

</details>

### Exercise 1.9.3

Use `nrow()`, `ncol()`, `names()`, and `glimpse()` on the built-in dataset `mtcars`. How many cars (rows) and variables (columns) are there? What types are the columns?

<details> <summary>Solution</summary>

```{r}
nrow(mtcars)
ncol(mtcars)
names(mtcars)
glimpse(mtcars)
```

There are 32 cars and 11 variables. All columns are `<dbl>` (numeric/double).

</details>

### Exercise 1.9.4

Using bracket notation `[row, column]` on `mtcars`, extract:

1. The value in row 3, column 1
2. All of row 1
3. All rows where `mpg` is greater than 30

<details> <summary>Solution</summary>

```{r}
mtcars[3, 1]
mtcars[1, ]
mtcars[mtcars$mpg > 30, ]
```

</details>

### Exercise 1.9.5

There is a subtle but important difference between `mtcars["mpg"]` and `mtcars[["mpg"]]` (single vs. double brackets). Run both and check the result with `class()`. What type does each return?

<details> <summary>Solution</summary>

```{r}
class(mtcars["mpg"])
class(mtcars[["mpg"]])
```

Single brackets `["mpg"]` return a **data frame** (a table with one column). Double brackets `[["mpg"]]` return the column as a **vector**. The `$` operator (`mtcars$mpg`) behaves like double brackets — it extracts the vector. This distinction matters when you need a vector (for arithmetic, passing to functions) versus when you want to keep the data frame structure.

</details>