## Indexing and subsetting

{{< video https://vimeo.com/1198722235/140449fd65?fl=ml&fe=ec >}}

We learn how to extract specific elements from a vector using integer indices and logical conditions, which is the basis for all subsetting in R.


```{r, eval = FALSE}
prices   <- c(29.90, 35, 11, 12.5, 48, 22.75)
product  <- c("coffee", "tea", "juice", "sandwich", "cookie")

# Indexing a vector
prices[3]                    # By position
prices[c(1, 2, 5)]           # By several positions
prices[2:4]                  # Using a sequence
prices[-1]                   # Excluding elements

# Using logical vectors, and thus conditions
prices[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
prices[prices > 30]
products[prices > 30]

# Using indexing to change specific elements
prices[3] <- 15
prices[prices > 30] <- prices[prices > 30] * 0.8 
```

## Exercises

### Exercise 1.8.1

Using the vector:

```r
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
```

Extract the third element. Extract the last three elements.

<details> <summary>Solution</summary>

```{r}
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

months[3]
months[10:12]
```

Alternatively for the last three: `tail(months, 3)`.

</details>

### Exercise 1.8.2

Using the vector `sales <- c(120, 85, 200, 340, 95, 180, 420, 60, 275, 150)`, extract all values that are between 100 and 300 (inclusive).

<details> <summary>Solution</summary>

```{r}
sales <- c(120, 85, 200, 340, 95, 180, 420, 60, 275, 150)

# We can do this in two steps with what we have learned so far:
sales2 <- sales[sales >= 100]
sales2[sales2 <= 300]

# But it is easier to use a logical "AND", which we will learn in the next topic:
sales[sales >= 100 & sales <= 300]
```
</details>

### Exercise 1.8.3

Replace all values in `sales` that are below 100 with 100 (a "floor" operation). Print the result.

<details> <summary>Solution</summary>

```r
sales[sales < 100] <- 100
sales
```

```
[1] 120 100 200 340 100 180 420 100 275 150
```

Alternatively, you can use `pmax(sales, 100)` which returns the element-wise maximum of `sales` and 100.

</details>

### Exercise 1.8.4

Given:

```r
x <- c(5, 3, 8, 1, 9, 2, 7)
```

Use `sort()` and `order()` on `x`. What does each return? How are they different?

<details> <summary>Solution</summary>

```{r}
x <- c(5, 3, 8, 1, 9, 2, 7)

sort(x)
order(x)
```

`sort(x)` returns the values in sorted order. `order(x)` returns the _positions_ that would sort the vector: the 4th element (1) is smallest, the 6th element (2) is next, etc.

You can use `order()` to sort one vector by another: `x[order(x)]` is the same as `sort(x)`.

</details>

### Exercise 1.8.5

What does the following code return? Think carefully before running it.

```{r}
x <- 1:10
x[-c(3, 5, 7)] * 2
```

<details> <summary>Solution</summary>

```{r}
x <- 1:10
x[-c(3, 5, 7)] * 2

`x[-c(3, 5, 7)]` removes elements at positions 3, 5, and 7, giving `c(1, 2, 4, 6, 8, 9, 10)`. Multiplying by 2 gives `c(2, 4, 8, 12, 16, 18, 20)`.

</details>

