## Basic data frame manipulation and the pipe

{{< video https://vimeo.com/1198722461/cd2f768130?fl=ml&fe=ec >}}

We introduce the native pipe operator `|>`, which lets you chain operations together in a readable left-to-right sequence, and apply it to a small end-to-end example.


```{r, eval = FALSE}
library(tidyverse)

prices <- c(29.90, 35, 11, 12.5, 48, 22.75)
products <- c("coffee", "tea", "juice", "sandwich", "cookie")
quantities <- c(3, 5, 0, 2, 8)

sales <- tibble(
  product = products,
  price = prices,
  quantity = quantities
)

# Creating a new column
sales$revenue <- sales$price * sales$quantity

# The pipe
head(sales, n = 2)        # Without the pipe
sales |> head(n = 2)      # The same operation, but with the pipe

# Using the pipe for data frame operations
sales |> 
  head(n = 2) |> 
  nrow()

# Quick demonstration, requires the tidyverse package to be loaded.
sales |> 
  filter(quantity > 0) |> 
  arrange(revenue)
```

## Exercises

### Exercise 1.10.1

Rewrite the following using the pipe `|>`:

```r
head(mtcars, 3)
```

<details> <summary>Solution</summary>

```{r}
mtcars |> head(3)
```

</details>

### Exercise 1.10.2

Use the pipe to chain `summary()` on the `mtcars` dataset:

```r
mtcars |> summary()
```

Now try piping `mtcars` through `colMeans()`. Does it work? Why or why not?

<details> <summary>Solution</summary>

```{r}
mtcars |> summary()

mtcars |> colMeans()
```

Both work because `summary()` and `colMeans()` take a data frame as their first argument, which is what the pipe provides. If a function does not take a data frame as its first argument, the pipe will not work as smoothly.

</details>

### Exercise 1.10.3

Chain three operations with pipes: take `mtcars`, extract just the `mpg` column with `$`, then compute the `range()`:

```r
mtcars$mpg |> range()
```

Now write the same thing without the pipe. Which version do you find more readable?

<details> <summary>Solution</summary>

```{r}
# With pipe
mtcars$mpg |> range()

# Without pipe
range(mtcars$mpg)
```

For a single function call, the non-pipe version is arguably simpler. The pipe becomes more useful when chaining two or more operations together. There is no need to force the pipe when a simple function call is clearer.

</details>

### Exercise 1.10.4

Examine the difference between the two pipe operators. Run:

```r
mtcars |> head(3)
```

Now load the magrittr-package (it is installed as part of the tidyverse) and try:

```r
library(magrittr)
mtcars %>% head(3)
```

Both should give the same result. When you look up R code online (Stack Overflow, blog posts, etc.), which pipe are you more likely to see? Why?

<details> <summary>Solution</summary>

Both give identical results:

```
               mpg cyl disp  hp drat    wt  qsec vs am gear carb
Mazda RX4     21.0   6  160 110 3.90 2.620 16.46  0  1    4    4
Mazda RX4 Wag 21.0   6  160 110 3.90 2.875 17.02  0  1    4    4
Datsun 710    22.8   4  108  93 3.85 2.320 18.61  1  1    4    1
```

You will see `%>%` far more often online, because it has been available since 2014 (via the magrittr package), while `|>` was only introduced in R 4.1 (2021). Most existing tutorials, Stack Overflow answers, and blog posts were written before the native pipe existed. Both work essentially the same way for our purposes. We use `|>` in this course because it requires no extra packages and is the modern standard.

</details>
